---
title: "R Notebook"
output: html_notebook
---

## Double check we are in the correct directory

```{r}
main_dir = "~/Desktop/af-werx"
data_dir = paste0(main_dir, "/data")
data_dir

setwd(data_dir)
getwd() #double check
```

## Load Data 
```{r}
load("tidyr_tables.RData") #load data
flights =  nycflights13::flights #assign data to variable

```

## Manipulating Data - Arrange Function
```{r}
arrange(flights,year,month,day) #arrange data (flights) by year, then month, then day (all start at 1)

arrange(flights,year, desc(month), day) #starts the months at 12 and counts down to 1 

NA_df = data.frame(x=c(1,NA,2), y=c(1,2,3)) #sample matrix
arrange(NA_df, desc(x)) #NAs go to bottom-- NA_df data is being organized according to x, in descending order

```

## Manipulating Data - Select Function
```{r}
#select only year month and day
select(flights, year, month, day)

#can also use the range operator 
select(flights, year:day)

#to exclude
select(flights, -(year:day))

#helper functions-- suppose we want the columns that end with delay
select(flights, ends_with("delay"))
```

## Manipulating Data -- Mutating and Transmuting Columns
```{r}

## Create a subset of the dataset
flights_sml = select(flights, year:day, ends_with("delay"), distance, air_time)
flights_sml

# Mutate -- lets add two columns called gain and speed
mutate(flights_sml, gain=arr_delay-dep_delay, speed=(distance/air_time)*60)

# Transmute -- if we only want the new columns we can use transmute
transmute(flights_sml, gain=arr_delay-dep_delay, speed=(distance/air_time)*60)

transmute(flights_sml, arr_delay, lag(arr_delay)) #shifts values down by one row
transmute(flights_sml, arr_delay, lead(arr_delay)) #shifts values up by one row
transmute(flights_sml, arr_delay, cumsum(arr_delay)) #cumulatively sums all values

```

## Manipulating Data - Rankings
```{r}
x <- c(5, 1, 3, 2, 2, NA)
row_number(x)
min_rank(x)
dense_rank(x)
percent_rank(x)
cume_dist(x)

```

## Summarise and Group_By
```{r}
# Summarise -- doesn't do much on its own. The function structure (what goes inside the ()) is whats important
# for example, summarise will output one value (mean result) from the flights table, from the dep_delay column
summarise(flights, delay=mean(dep_delay, na.rm=T))


# Group By -- starts delineating between groups of info
by_day = group_by(flights, year, month, day)
by_day

# Use Summarise and Group By Together -- Here's the bread and butter!!! 
summarise(by_day, delay=mean(dep_delay, na.rm=T)) 
#here summarise will output one value (mean result) for the dep delays, per day, per month, per year, from flights table!!! so each day will have its own summarise'd value (result)

# Group Flights by Destination
by_dest = group_by(flights, dest)

delay = summarise(by_dest, count=n(), 
                            dist=mean(distance, na.rm=T),
                            delay=mean(arr_delay, na.rm=T))
#here we find the mean arrive delay and mean distance for each destination! each destination gets its own summarise'd result 
                  
delay = filter(delay, count>20, dest!= "HNL")

delay


# Do the same as Summarise and Group By, but with Pipes
delays = flights %>% group_by(dest) %>% 
  summarise(count=n(), dist=mean(distance, na.rm=T), delay=mean(arr_delay, na.rm=T))
delays
```

### Another Summarise Example (2)
```{r}
# Use another example
flights %>% group_by(year, month, day) %>% 
  summarise(mean=mean(dep_delay, na.rm=T))

delays = flights %>% group_by(dest) %>% 
  summarise(count = n_distinct(distance), 
            dist=mean(distance, na.rm = T), 
            delay = mean(arr_delay, na.rm=T))
delays

# Use n() to count 
flights %>% group_by(year, month, day) %>% 
  summarise(mean=mean(dep_delay, na.rm =T), n = n())

flights %>% count(day)

#summarise with ranking
flights %>% group_by(year, month) %>% 
  summarise(first=min(dep_time, na.rm=T), last=max(dep_time, na.rm=T))

#just a demo
min(flights$dep_time, na.rm=T))
max(flights$dep_time, na.rm=T))
```
### Another Summarise Example (3)
```{r}

# Summarise Position
# not canceled flights

not_cancelled = flights %>% filter(!is.na(dep_time)) #remvoe canceled flights, where dep time is NA

not_cancelled %>% group_by(year, month, day) %>% summarise(first=min(dep_time), 
                                                           last = max(dep_time))

not_cancelled %>% group_by(year, month, day) %>% summarise (flights_that_take_off = n_distinct(dep_time)) #not best use of n distinct, but it works

```

## Ungrouping 
```{r}
# Ungrouping 
notcancelled %>% 
  ungroup() %>%   #more like regrouping, as opposed to ungrouping 
  group_by(year, month) %>% 
  summarise(flights_by_year = n_distinct(dep_time))
```

## Data Wrangling Intro -- Gathering and Spreading Columns
```{r}
#gather function, by name -- when data is wide and we want long (too many columns that could be used as rows)
year_country %>% 
  gather('1999','2000', key="year",value="cases")

# by range
year_country %>% 
  gather(2:3, key="year",value="cases")


# Spread -- when something is account for in the rows that should be its own variable (column)
key_value_country 

key_value_country %>% spread(key=key, value=value) #address by pipline and spread()

spread(key_value_country, key=key, value=value) #address with just spread()

```

## Data Wrangling -- Uniting and Separating Data
```{r}

# if we want to separate values
# 745/19987071 ---> 745 and 19987071
# we can use separate

rate_country

rate_country %>% 
  separate(rate, into=c("cases","population", sep="/"))

## Another example

ex_table = rate_country %>% 
  separate(year, into=c("century","year"), sep=2, convert=T) #sep=2 means separate after the second character
#convert changes the value into what R thinks is the best format (character to integer)

# Unite Function
ex_table %>% 
  unite(time, century, year, sep="") #separator is nothing (should unite clean), otherwise underscore is default


```
