---
title: "Intro to R"
output: html_notebook
---

## Setup for R Notebooks 
```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir="~/Desktop/af-werx/data") 

```

## Setting the Directory 
```{r}
setwd("~/Desktop/af-werx")
```


## Setting Working Directory 
```{r}
getwd()
```

## Set Our Directories
```{r}
main_dir = "~/Desktop/af-werx"
data_dir = paste0(main_dir, "/data")
data_dir
plot_dir = paste0(main_dir, "/plots")
plot_dir
setwd(data_dir)
getwd()
```
```{r}
temp_heart_rate = read.csv("temp_heart_rate.csv", header = T, stringsAsFactors = F)
```

## Structure of the Data
```{r}
str(temp_heart_rate)
```

## View First Row of Data 
```{r}
head(temp_heart_rate)
```

## View Last Rows of Data
```{r}
tail(temp_heart_rate)
```

## View the Whole Data Set
(Opens a new window)
```{r}
View(temp_heart_rate) #must be uppercase
```

## Create a Subset from Data (Writing Data)
```{r}
temp_heart_subset = temp_heart_rate[1:10,] #dont forget comma
write.csv(temp_heart_subset, "temp_heart_subset.csv", row.names = F)
```

## View (list) Objects in the Environment
```{r}
ls()
```

## Save Environment 
```{r}
save.image(file = "myenv.RData")
```

## Remove Variable
```{r}
rm("temp_heart_rate")
ls()
```

## Remove Whole Environment 
```{r}
rm(list=ls()) #everything is gone, variables, and directories 
```

## Load the Environment
```{r}
load("myenv.RData")
```


# Work Example - CMP Dataset
```{r}
CMP = read.csv("ChemicalManufacturingProcess.csv", header = T, stringsAsFactors = F)
str(CMP)
```

## View CMP Data
```{r}
View(CMP)
```

## Create a Subset of Data
```{r}
col_ids = c(1:4, 14:16)
col_ids

CMPsubset = CMP[,col_ids]
str(CMPsubset)

```


## Obtain Summary of Subset
```{r}
summary(CMPsubset)
```


## Manually Obtain Max Value of Column (Attribute)
```{r}
# Example with error
# maxprocess01 = max(CMPsubset$ManufacturingProcess01)
# maxprocess01 ##get NA because there is an NA within the data. Thus, cannot get max

#Fix
maxprocess01 = max(CMPsubset$ManufacturingProcess01, na.rm = T) #ignores NAs, does not remove NAs
maxprocess01 

```

## Identifying NAs within Columns (Attributes)
```{r}
is_na = is.na(CMPsubset$ManufacturingProcess01) #first value is NA, all else are good; dont need variable but makes it easier to ID with the which command

which(is_na) #easy shortcut
# equivalent to which(is.na(CMPsubset$ManufacturingProcess01))

CMPsubset$ManufacturingProcess01[is_na] #finds the NA! COOL

```

## Manually Obtain the Mean of a Column (Attribute)
```{r}
mean_process01 = mean(CMPsubset$ManufacturingProcess01, na.rm = T)
mean_process01
```

## Replace NAs with the Mean
```{r}
CMPsubset$ManufacturingProcess01[is_na] = mean_process01
# NA is being found, and set equal to Mean
```

## Recalculating the Max (with NAs replaced by Mean)
```{r}
maxprocess01 = max(CMPsubset$ManufacturingProcess01)
maxprocess01
```

# Repeat Examples -  Manufacturing Process 02
```{r}
mean_process02 = mean(CMPsubset$ManufacturingProcess02, na.rm = T) #Find Mean
mean_process02

is_na = is.na(CMPsubset$ManufacturingProcess02) #Find NAs

CMPsubset$ManufacturingProcess02[is_na] = mean_process02 #Replace NAs with Mean

maxprocess02 = max(CMPsubset$ManufacturingProcess02) #ReRun Max
maxprocess02
```
## Repeat Examples  - Manufacturing Process 03
```{r}
mean_process03 = mean(CMPsubset$ManufacturingProcess03, na.rm = T) 
mean_process03

is_na = is.na(CMPsubset$ManufacturingProcess03) 

CMPsubset$ManufacturingProcess03[is_na] = mean_process03

maxprocess03 = max(CMPsubset$ManufacturingProcess03)
maxprocess03
```

# Control Structure (If Else Statements)

## Obtain the Mean of Yield
```{r}
yield_mean = mean(CMP$Yield)
yield_mean
```

## Create New Column (Conditioned to Abv Avg/Bel Avg Yield)
```{r}
#col created/named    take OG    compare mean   success          failure
CMP$new_yield = ifelse(CMP$Yield >= yield_mean, "above_average", "below_average") #creating and naming a new column within master file, column is conditioned: if original yield is greater than or equal to yield mean, succes = above avg, fail = below avg
head(CMP[,c("Yield","new_yield")]) #view head of results, vector of original yield and new yield 
tail(CMP[,c("Yield","new_yield")]) #view tail of results, being the original yield and new yield
write.csv(CMP[,c("Yield","new_yield")], "newyield.csv", row.names=F) #write results to new CSV file

View(CMP)
```

## For Loops
```{r}
CMPsubset_variables = colnames(CMPsubset)

start = 3
end = 6

for (i in start:end){
  print(CMPsubset_variables[i])
}

```
## Make a Function
```{r}
PrintHello = function(name){
  hello = paste0("hello ", name)
  print(hello)
}

PrintHello("AFWERX NEXUS")
```
## Make a Better Function
```{r}
GetPi = function(n=4){
  pi = round(3.141592,n)
  return(pi)
}

GetPi(5)
```
## Create a Function That Can Impute NAs with Mean
```{r}
ImputeNAsWithMean = function(dataset){
  for(i in 1:ncol(dataset)){
    is_na = is.na(dataset[,i])
    if(any(is_na)){  ##will ring true if equals true, triggered by previous command which results in logical value (true or false)
      na_ids = which(is_na) 
      
      var_mean = mean(dataset[,i], na.rm = T)
      dataset[na_ids,i] = var_mean
      message = paste0("Replacing NA with mean of: ", colnames(dataset)[i])
      print(message)
    }
  }
  return(dataset)
}

CMPsubset = CMP[,c(1:4, 14:16)]
CMPsubset_imputed = ImputeNAsWithMean(CMPsubset)
View(CMPsubset_imputed)

```

## Calling Functions
```{r}
source("ImputeNAsWithMean.R") #careful on where they are saved, should be in the same directory as the dataset unless specifically defined

GetPi(3)
 
```
