---
title: "R Notebook"
output: html_notebook
---


## Getting the Data
```{r}
# View State.x77
View(state.x77)


class(state.x77) #its a matrix 
str(state.x77)


dim(state.x77) #get dim dims
dim(state.x77)[1] # outputs number of rows 
dim(state.x77)[2] # outputs number of columns

```

## Getting a Random Sample
```{r}
# Getting Samples
set.seed(1) #running the sample lines alone produces new results; if you want same results, run seed first every time
sample(1:100, size=10)
sample(state.name, size = 5)
```

## Defining Functionals (lapply)
```{r}
## For Loops and lapply
out_list = list()
 
for(i in 1:15){
  out_list[i] = sqrt(i)
}

out_list # below line of code is the exact same thing as this

#---

out_list = lapply(1:5, sqrt) # samesies 
out_list
```

##  Unique Functionals -- removes repeated values and compiles unique values, using lapply
```{r}

state.region #auto loaded
unique_regions = unique(state.region)
unique_regions

unique_regions_list = as.list(unique_regions) #convert to list, remember, not same as vector, values have "embedded titles"
unique_regions_list

as.character(unique_regions_list[[1]]) ##double brackets for characters (factors with levels), or else get level

unique_regions_list = lapply(unique_regions_list, as.character) ##get as character not factor 
unique_regions_list

toupper(unique_regions_list[[1]]) ##convert to uppercase

unique_regions_list = lapply(unique_regions_list, toupper) #convert to uppercase
unique_regions_list
```

## Create a Numeric List, Obtain Absolute Values, Using lapply
```{r}
#create a numeric list
numeric_list = list(first=-1, second=2, third=3, fourth=-4)
numeric_list_abs = lapply(numeric_list, abs) #suppose we want only positive numbers
numeric_list_abs #we can apply the abs() function to the list to remove negatives
```

## GSUB -- Substitute Values within a data set
```{r}
gsub(" ","_", unique_regions_list[[4]]) #first is replacement value, second is replacement, third is target

#new list    universal change to reg list        custom function, replace spaces with underscore, target is the uppercase version of the target
fancy_list = lapply(unique_regions_list, function(entry){gsub(" ","_",toupper(entry))})
fancy_list
```

## Creating a function using GSUB -- replace spaces with underscores and uppercase 
```{r}
# lets create a function with a name to use it with lapply 
toUpperAndUnderscore = function(entry){
  gsub(" ","_",toupper(entry))
} # Boom, got a function that replaces spaces with underscores, and then applies it to the uppercase target

toUpperAndUnderscore("north central") #test that it works (it does)

fancy_list = lapply(unique_regions_list, toUpperAndUnderscore) #run an application loop where UAU is applied to URL
fancy_list ##no "()" entry placeholder needed for UAU within lapply
#lapply takes values one by one (application loop), so stick with lists and vectors

```

## NEW EXAMPLE -- Generate a Random Sequence of Numbers using Sample
```{r}
# Generate a Random Sequence of Numbers
sample_vec = sample(1:50, size = 50)
sample_vec #got a random vector of 1-50

state.name
state.name[sample_vec] #shuffles deck of 50 state names
```

## Create a function that compiles a list of even States (according to their randomized identifiers) (creating a random sample group)
```{r}
if(sample_vec[5] %% 2 == 0){ ##checks if value is even; if remainder after dividing by 2 is 0
  TRUE #then output true
} else{
  FALSE #otherwise output false
}


#lets create a function that returns TRUE if value is EVEN
IsEven = function(n){
  if(n %% 2 == 0){
    TRUE
  } else{
    FALSE}
}

logical_list = lapply(sample_vec, IsEven) ##only works with lists and vectors, produces list
logical_list #created a logical list (true, evens) corresponding to state shuffle

#we can make it into a vector or flatten it with unlist()
logical_vec = unlist(logical_list)
logical_vec

#we want even ids so we can get the names of the states
even_ids = sample_vec[logical_vec] #takes list of 50 random numbers 1-50, and only takes true values (evens)
even_ids 

state.name[even_ids] #produce state names according to randomized number list (evens only)


```

## Create a function that compiles a list of Odd States
```{r}

#lets create a function that returns TRUE if value is ODD
IsOdd = function(n){
  if(n %% 2 != 0){
    TRUE
  } else{
    FALSE}
}

logical_list2 = lapply(sample_vec, IsOdd)
logical_list2 #created a logical list (true, evens) corresponding to state shuffle

#we can make it into a vector or flatten it with unlist()
logical_vec2 = unlist(logical_list2)
logical_vec2

#we want even ids so we can get the names of the states
odd_ids = sample_vec[logical_vec2] #takes list of 50 random numbers 1-50, and only takes true values (evens)
odd_ids 

state.name[odd_ids] #produce state names according to randomized number list (evens only)


```

## SAPPLY -- Loop based command that applies a function to each individual value 
```{r}

# SAPPLY 
logical_vec = sapply(sample_vec, IsEven) ##outputs as a vector, can handle dataframes, lists, and vectors
logical_vec

state_data = state.x77 
state_df =  as.data.frame(state_data) #converts state_data (a matrix) to a dataframe
class(state_data) #this is a matrix
class(state_df) #this is a dataframe

rownames(state_df) #gets row names

state_df$State = rownames(state_df) #Row names are states, should now be in a new row title column
View(state_df)

rownames(state_df) = NULL #to revmove row names we just created
View(state_df)

variable_class = sapply(state_df, class) #produces vector of the class of each variable
variable_class

variable_type = sapply(state_df, typeof) #prdouces vector of type of each variable
variable_type

```


## Installing Packages and Loading Libraries
```{r}

## Libraries and Packages

library(tidyverse) #import a library
library(help = "tidyverse") # help with the specific library

# sometimes you may see required instead of a library but it also loads it
require(tidyverse) #samesies 

install.packages("nycflights13") #install package
library(nycflights13) #load corresponding library (libraries are bigger than packages)
```


## Loading a Table from a Library
```{r}
getwd() ##double check current working directory (should be desktop, afwerx, data)
load("tidyr_tables.RData") #load the tables we want from the directory

# load nyc flight dataset
flights = nycflights13::flights #take table named flights from the nycflights13 library, assign it to variable named flights
View(flights)

#load data into environment 
data(flights)
filtered_flights = filter(flights, month == 1, year == 2013) #filter flights according to month and year
```

## Filtering and Manipulating a Table
```{r}

#month 1 and day 25, and dep time 15
filter(flights, month == 1 & day == 25 & dep_time==15)

#month 1, 3, and 8, and day 25
filter(flights, month %in% c(1, 3, 8) & day == 25)

#filter with NA
NA_df = data.frame(x = c(1, NA, 2), y = c(1,2,3)) #create NA_df dataframe (1 NA 2; 1 2 3)

filter(NA_df, x>=1) #filter NA_df and return everything that meets criteria (x>=1)

filter(NA_df, is.na(x) | x>=1) #filter NA_df, return x>=1 OR NAs --- you KEEP what meets criteria 

```
